---
title: "ridgereg"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ridgereg}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


#' ---
#' title: "Using ridgereg for Ridge Regression and Model Comparison"
#' ---

## Introduction

In this vignette, we demonstrate how to use the `ridgereg()` function implemented
in our package to perform ridge regression and compare its performance to
ordinary least squares regression using the `caret` package.

```{r}
# the needed packages
library(mlbench)
library(caret)
library(leaps)
library(dplyr)
library(ggplot2)
library(Lab4)

data("BostonHousing")

```



Split data into training-data and test-data:

```{r}
set.seed(42)
train_index <- createDataPartition(BostonHousing$medv, p = 0.8, list = FALSE)
train_data <- BostonHousing[train_index, ]
test_data  <- BostonHousing[-train_index, ]

```



Now we do linear regression using caret:

```{r}
# a simple linear regression using caret
lm_full <- train(
  medv ~ ., 
  data = train_data,
  method = "lm",
  trControl = trainControl(method = "none") 
)

# and a forward selection model
lm_forward <- train(
  medv ~ ., 
  data = train_data,
  method = "leapForward",
  tuneGrid = data.frame(nvmax = 1:8), 
  trControl = trainControl(method = "cv", number = 5)
)

summary(lm_forward$finalModel)


```

This confirms that the forward selection algorithm trained one model at each subset size (1, 2, 3, 4, 5, 6)

Now we pick out the best forward selection model and evaluate:

```{r}

# get the best model out
best_vars <- names(coef(lm_forward$finalModel, 6))[-1]
formula_forward <- as.formula(paste("medv ~", paste(best_vars, collapse = " + ")))

# refit as a standard lm model
lm_forward_best <- lm(formula_forward, data = train_data)

# evauate on train-data
pred_train_forward <- stats::predict(lm_forward_best, newdata = train_data)
rmse_forward_train <- sqrt(mean((train_data$medv - pred_train_forward)^2))
r2_forward_train <- 1 - sum((train_data$medv - pred_train_forward)^2) / 
                          sum((train_data$medv - mean(train_data$medv))^2)

data.frame(
  Model = "Forward Selection (6 predictors)",
  RMSE = rmse_forward_train,
  R2 = r2_forward_train
)
```

The model explains about 71.4% of the variance in median home values in the training data and the predictions differ from the actual house values (*medv*) by about 4.83 units (thousands of dollars).

Now we fit some ride regression models with *ridgereg()* from our package **Lab4**. Testing some diffrent $\lambda$ values.

```{r}
# ridge regression using ridgereg() from our Lab4 package
ridge1 <- ridgereg(medv ~ ., data = train_data, lambda = 0.1)
ridge2 <- ridgereg(medv ~ ., data = train_data, lambda = 1)
ridge3 <- ridgereg(medv ~ ., data = train_data, lambda = 10)
```




















