---
title: "ridgereg"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ridgereg}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  fig.align = "center",
  dpi = 150
)
```




**Introduction**

In this vignette, we demonstrate how to use the `ridgereg()` function implemented
in our package to perform ridge regression and compare its performance to
ordinary least squares regression using the `caret` package. This vignette also 
shows **mean_flight_delays** trough the function *visualize_airport_delays()*.

```{r}
# the needed packages
library(mlbench)
library(caret)
library(leaps)
library(dplyr)
library(ggplot2)
library(nycflights13)
library(Lab4)

data("BostonHousing")

```


# ridgereg

Split data into training-data and test-data:

```{r}
set.seed(42)
train_index <- createDataPartition(BostonHousing$medv, p = 0.8, list = FALSE)
train_data <- BostonHousing[train_index, ]
test_data  <- BostonHousing[-train_index, ]

```



Now we do linear regression using caret:

```{r}
# a simple linear regression using caret
lm_full <- train(
  medv ~ ., 
  data = train_data,
  method = "lm",
  trControl = trainControl(method = "none") 
)

# and a forward selection model
lm_forward <- train(
  medv ~ ., 
  data = train_data,
  method = "leapForward",
  tuneGrid = data.frame(nvmax = 1:8), 
  trControl = trainControl(method = "cv", number = 5)
)

summary(lm_forward$finalModel)


```

This confirms that the forward selection algorithm trained one model at each subset size (1, 2, 3, 4, 5, 6)

Now we pick out the best forward selection model and evaluate:

```{r}

# get the best model out
best_vars <- names(coef(lm_forward$finalModel, 6))[-1]
formula_forward <- as.formula(paste("medv ~", paste(best_vars, collapse = " + ")))

# refit as a standard lm model
lm_forward_best <- lm(formula_forward, data = train_data)

# evauate on train-data
pred_train_forward <- stats::predict(lm_forward_best, newdata = train_data)
rmse_forward_train <- sqrt(mean((train_data$medv - pred_train_forward)^2))
r2_forward_train <- 1 - sum((train_data$medv - pred_train_forward)^2) / 
                          sum((train_data$medv - mean(train_data$medv))^2)

data.frame(
  Model = "Forward Selection (6 predictors)",
  RMSE = rmse_forward_train,
  R2 = r2_forward_train
)
```

The model explains about 71.4% of the variance in median home values in the training data and the predictions differ from the actual house values (*medv*) by about 4.83 units (thousands of dollars).

Now we fit some ride regression models with *ridgereg()* from our package **Lab4**. Testing some diffrent $\lambda$ values.

```{r}

set.seed(42)

lambdas <- c(0.01, 0.1, 1, 10)

# 10 folds
k <- 10
folds <- sample(rep(1:k, length.out = nrow(train_data)))

cv_results <- data.frame(lambda = lambdas, RMSE = NA)

# cross-validation
for (i in seq_along(lambdas)) {
  lambda_i <- lambdas[i]
  fold_rmse <- numeric(k)
  
  for (fold in 1:k) {
    train_fold <- train_data[folds != fold, ]
    valid_fold <- train_data[folds == fold, ]
    

    model_i <- ridgereg(medv ~ ., data = train_fold, lambda = lambda_i)
    
    # preds
    preds <- predict(model_i, newdata = valid_fold)
    
    # RMSE for this fold
    rmse <- sqrt(mean((valid_fold$medv - preds)^2))
    fold_rmse[fold] <- rmse
  }
  
  cv_results$RMSE[i] <- mean(fold_rmse)
}

print(cv_results)
best_lambda <- cv_results$lambda[which.min(cv_results$RMSE)]
cat("Best lambda:", best_lambda, "\n")

# save it
ridge_best <- ridgereg(medv ~ ., data = train_data, lambda = best_lambda)

```




Now we compare the three models:

```{r}

rmse <- function(actual, predicted) {
  sqrt(mean((actual - predicted)^2))
}


pred_full <- predict(lm_full, newdata = test_data)
pred_forward <- predict(lm_forward_best, newdata = test_data)
pred_ridge <- predict(ridge_best, newdata = test_data)

r2 <- function(actual, predicted) {
  cor(actual, predicted)^2
}


results <- data.frame(
  Model = c("Full Linear Model", "Forward Selection", "Ridge Regression"),
  RMSE = c(
    rmse(test_data$medv, pred_full),
    rmse(test_data$medv, pred_forward),
    rmse(test_data$medv, pred_ridge)
  ),
  R2 = c(
    r2(test_data$medv, pred_full),
    r2(test_data$medv, pred_forward),
    r2(test_data$medv, pred_ridge)
  )
)

print(results)

```


The Ridge regression with $\lambda$ = 1 is the best choice for our data. It has comparable predictive performance to the full OLS model while offering better coefficient stability and generalization.



# Mean_flight_delays

To show the function *visualize_airport_delays()* we just run the function:

```{r}

visualize_airport_delays()
```

The function creates an **ggplot2** output where each point represents an airport’s 
location based on latitude and longitude. The color scale shows the average arrival 
delay in minutes, warmer colors indicate longer delays, while cooler colors indicate 
shorter or early arrivals.




